Thermal reactionomes reveal divergent responses to thermal extremes in warm

### 2015-06-03 ApTranscriptome_TR.Rmd
revisions for MolEco

mh_plot_sub1 <- qplot(y=Apic, x=Acar, data=md_sub1, fill=value, geom="tile", xlim = rev(levels(md_sub1$Acar))) + 
  scale_fill_gradient2(limits=c(-700, 600), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. picea", x = "A. carolinensis") +
  theme(legend.position = "none") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

@@

mh_plot_sub2 <- qplot(y=Acar, x=Apic, data=md_sub2, fill=value, geom="tile", xlim = rev(levels(md_sub2$Apic))) + 
  scale_fill_gradient2(limits=c(-700, 600), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. carolinensis", x = "A. picea") + 
  theme(legend.position = "right") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 2.5, ymax = 3.5, fill = "transparent", linetype=2, colour="black")

@@

png("results/matched_observations_Fig2.png")
grid.arrange(mh_plot_sub1, mh_plot_sub2, ncol=2)
dev.off()

@@

# plot
enhanced_defense_plot <- qplot(y=Apic, x=Acar, data=enhanced_defense_dat, fill=value, geom="tile", xlim = rev(levels(enhanced_defense_dat$Acar))) + 
  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. picea", x = "A. carolinensis") +
  theme(legend.position = "none") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

# title
enhanced_defense_plot <- enhanced_defense_plot + ggtitle('Enhanced Response hypothesis') + theme(plot.title = element_text(hjust=0))
enhanced_defense_plot
```

@@

# plot
tolerance_plot <- qplot(y=Apic, x=Acar, data=tolerance_dat, fill=value, geom="tile", xlim = rev(levels(tolerance_dat$Acar))) + 
  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. picea", x = "A. carolinensis") +
  theme(legend.position = "bottom") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

# title
tolerance_plot <- tolerance_plot + ggtitle('Tolerance hypothesis') + theme(plot.title = element_text(hjust=0))
tolerance_plot

# plot together
png("results/predictions_Fig1.png")
grid.arrange(enhanced_defense_plot, tolerance_plot, ncol=2)
dev.off()
```

@@

```{r shiny_file, eval=FALSE, echo=FALSE}
shiny.dt <- TPM.dt.sub[responsive.lms.ann.type[,list(Transcript, best.hit.to.nr.1, GO.Biological.Process.1)]]

>>>>

mh_plot_sub1 <- qplot(y=Apic, x=Acar, data=md_sub1, fill=value, geom="tile", xlim = rev(levels(md_sub1$Acar))) + 
  scale_fill_gradient2(limits=c(-700, 600), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. picea", x = "A. carolinensis") +
  theme(legend.position = "right") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

@@

mh_plot_sub2 <- qplot(y=Acar, x=Apic, data=md_sub2, fill=value, geom="tile", xlim = rev(levels(md_sub2$Apic))) + 
  scale_fill_gradient2(limits=c(-700, 600), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. carolinensis", x = "A. picea") + 
  theme(legend.position = "none") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 2.5, ymax = 3.5, fill = "transparent", linetype=2, colour="black")

@@

png("results/matched_observations_Fig2.png")
grid.arrange(mh_plot_sub1, mh_plot_sub2, ncol=2, widths = c(3.75, 3))
dev.off()

@@

# plot
enhanced_defense_plot <- qplot(y=Apic, x=Acar, data=enhanced_defense_dat, fill=value, geom="tile", xlim = rev(levels(enhanced_defense_dat$Acar))) + 
  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. picea", x = "A. carolinensis") +
  theme(legend.position = "right") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

# title
enhanced_defense_plot <- enhanced_defense_plot + ggtitle('Enhanced Response hypothesis') + theme(plot.title = element_text(hjust=0, size = 12))
enhanced_defense_plot
```

@@

# plot
tolerance_plot <- qplot(y=Apic, x=Acar, data=tolerance_dat, fill=value, geom="tile", xlim = rev(levels(tolerance_dat$Acar))) + 
  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. picea", x = "A. carolinensis") +
  theme(legend.position = "none") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

# title
tolerance_plot <- tolerance_plot + ggtitle('Tolerance hypothesis') + theme(plot.title = element_text(hjust=0, size = 12))
tolerance_plot

# plot together
png("results/predictions_Fig1.png")
grid.arrange(enhanced_defense_plot, tolerance_plot, ncol=2, widths = c(3.75,3))
dev.off()
```

@@

```{r shiny_file, eval=FALSE, echo=FALSE}
shiny.dt <- TPM.dt.sub[responsive.lms.ann.type[,list(Transcript, best.hit.to.nr, GO.Biological.Process)]]

### 2015-06-03 doc/ApTranscriptome_manuscript.Rmd
revisions for MolEco

We quantified expression of each transcript using the program `Sailfish` [@patro2014] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of transcript expression. We included the contaminant transcripts identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses. As preliminary examination of the data (Supporting Information) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supporting Information).

To identify transcripts that have significant changes in expression across the thermal gradient, we fit the regression model

@@

@@

Though the total number of thermally-responsive transcripts did not differ between species ($\chi^2$~`r chi1$parameter`~ = `r round(chi1$statistic, 2)`, *P* = `r round(chi1$p.value, 2)`) there was significant qualitative and quantitative variation in thermal responses among transcripts and between the two species. To examine this, we grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C). The majority of the responsive genes were in the **Low** category for both species (42% and 54% for *A. carolinensis* and *A. picea*, respectively), followed by **Intermediate** (29%), **High** (12%),**Bimodal** (10%) and **Not Responsive** (8%) for *A. carolinensis*, and **Bimodal** (17%), **High** (14%), **Intermediate** (10%) and **Not Responsive** (6%) for *A. picea*.

@@

A number of caveats do apply to our work. First, species may differ in gene expression along axes which we have not measure here, primarily temporal patterns of gene expression [@sorensen2005], and could be studied in futher work. Second, the *de novo* transcriptome assembly is highly fragmented, given that all sequenced ant genomes to date have only about 18,000 genes [@gadau2012]. Although we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially assembled transcripts. A genome assembly, in progress, will help to reduce fragmentation. Third, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for the model organisms *Drosophila* and *Apis*. Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005], and large differences in patterns of gene expression may have only small effects on organismal fitness. In particular, functional protein levels cannot be expected to be directly linked to mRNA abundance due to post-transcriptional modification, regulation, mRNA fluctuations and protein stability [@feder2005].

>>>>

We quantified expression of each transcript using the program `Sailfish` [@patro2014] and used the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of transcript expression. We included the contaminant transcripts identified by `DeconSeq` at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses. As preliminary examination of the data (Supporting Information) indicated that the 7°C samples may have been mis-labeled, so we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supporting Information).

To identify transcripts that had significant changes in expression across the thermal gradient, we fit the regression model

@@

We identified thermally-responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC. All analyses were performed with R 3.1 [@r_core_team2013] and are fully reproducible (Supporting Information).

For each thermally-responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species. We used the predicted transcript expression levels to partition transcripts into five expression categories which were defined *a priori*, and could therefore be used to test predictions derived from three thermal adaptation hypotheses of relative response severity in the two species: **High** transcripts had greatest expression at temperatures > 31°C, **Low** transcripts had greatest expression at temperatures < 10°C, **Intermediate** transcripts had greatest expression between 10 to 30°C, **Bimodal** transcripts had increased expression at both high and low temperatures, while **NotResp** transcripts were those that were not thermally responsive in one species but did respond in the other. For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C. The two categories which unambiguously distinguish up- from down-regulation are *Bimodal* (up at both extremes) and *Intermediate* (down at both extremes).

@@

Though the total number of thermally-responsive transcripts did not differ between species ($\chi^2$~`r chi1$parameter`~ = `r round(chi1$statistic, 2)`, *P* = `r round(chi1$p.value, 2)`) there was significant qualitative and quantitative variation in thermal responses among transcripts and between the two species. The majority of the responsive genes were in the **Low** category for both species (42% and 54% for *A. carolinensis* and *A. picea*, respectively), followed by **Intermediate** (29%), **High** (12%),**Bimodal** (10%) and **Not Responsive** (8%) for *A. carolinensis*, and **Bimodal** (17%), **High** (14%), **Intermediate** (10%) and **Not Responsive** (6%) for *A. picea*.

@@

A number of caveats do apply to our work. First, species may differ in gene expression along axes which we have not measure here, primarily temporal patterns of gene expression [@sorensen2005], and could be studied in futher work. Second, the *de novo* transcriptome assembly is highly fragmented, given that all sequenced ant genomes to date have only about 18,000 genes [@gadau2012]. Although we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially assembled transcripts. A genome assembly, in progress, will help to reduce fragmentation. Third, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for the model organisms *Drosophila* and *Apis*. Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005], and large differences in patterns of gene expression may have only small effects on organismal fitness. In particular, functional protein levels cannot be expected to be fully linked to mRNA abundance due to post-transcriptional modification, regulation, mRNA fluctuations and protein stability [@feder2005].
